---
title: "Refugee Placements in the US, 2002-2019"
author: "Matt Brundage"
date: "May 8, 2019"
output: slidy_presentation
  
  
---

## Project goals

>- Visualize data on refugee placements in the U.S. by city, state, year, and refugee nationality (e.g., Iran, Syria, etc.)

>- Develop a function that takes user input (e.g., year(s) of interest, refugee placements as an absolute figure or as a percentage of the city population, etc.) and produces (1) a corresponding city "bubble" map and (2) a table of the top datapoints



## Data

>- Data are collected from the Refugee Processing Center (RPC) operated by the U.S Department of State Bureau of Population, Refugees, and Migration. You can explore the raw data here: http://ireports.wrapsnet.org/

>- The State Department FOIA request (submitted 3.5 years ago) for more detailed data on refugee placements remains in process.



## Data

>- The only other people (academics or otherwise) who seem to have made use of these data are a group of geographers based out of the University of Vermont. They have developed some of their own maps and have done qualitative work on the impact of placements in Vermont. See project details here: http://spatializingmigration.net/about/

>- Pablo Bose, the lead researcher on this project, received a $200,000 NSF grant in 2013 to conduct research on refugees in Vermont. See CV here: https://www.uvm.edu/sites/default/files/Bose_CV_Aug_11_2016.pdf 



## Data strengths

>- **Detailed.** Data are available for any selection of dates between 1/1/2002 and the present. Data are even available before 2002, but the RPC reports having less confidence in these figures. The cited reason is that pre-2002 data were migrated from a legacy system.

>- **Comprehensive.** Data include all placements of refugees of all nationalities since 1/1/2002. Data include 122 nationalities across 3,393 cities.



## Data weaknesses

>- **Demographics aside from nationality are not available by location.** That is, although the RPC offers data on religion, age, sex, gender, ethnicity, religion, language, and education by nationality for any set of dates (e.g., number of female Muslim arrivals from 1/1/2002 - 6/1/2002), it does *not* offer these data by location (e.g., number of female Muslim arrivals in Mobile, Alabama).

>- **Limited info on the substantive significance of "placements."** In particular, to what extent do placement data reflect the location where refugees ultimately live? Next steps include further digging into RPC data documentation, communication with folks at the RPC center, or communication with the folks working with the RPC data in Vermont.



## Data weaknesses

>- **Difficult to merge with relevant census data.** City data are matched to city populations provided by the U.S. Census "place" data, which provide the most granular measure of cities. The cities often span multiple county borders and all other potential U.S. Census borders of interest, so the "places" data are the only potential source of matching. Unfortunately, the RPC presents refugee data only by city name. There is no FIPS code, or anything analogous, which can be used to match the data to U.S. Census data. Often, the U.S. Census names do not perfectly match the RPC city names. Accordingly, the merging process is a bit messy and requires some judgment calls. There are still some cities without matches in the data, but I have taken care to ensure matches for the vast majority of cases.

>- **Takeway**: The data herein provide a very good sense of the magnitude of placements by geography, but should be interpreted with caution.



## Data visualization

>- **First**, I have included example city bubble maps and tables for several different visualizations of the data

>- For the city bubble maps, you can "hover" over cities to read their name and the associated outcome of interest (e.g., total arrivals from Iran)

>- **Second**, I include instructions for using my code, available on my github page, to play with the data yourself

>- The goal here is to be able to visualize any subset of interest within seconds



## Example 1

```{r include=FALSE}

# Setup ----

# Clear, set directory
rm(list=ls())
setwd("/Users/mattsair/Desktop/Projects/Refugee project/Data/refugee_data_clean")

refugee_data <- read.delim("refugee_data_clean.txt")

# Convert objects to their appropriate class
refugee_data$state <- as.character(refugee_data$state)
refugee_data$city <- as.character(refugee_data$city)
refugee_data$year <- as.numeric(refugee_data$year)
refugee_data$nationality <- as.character(refugee_data$nationality)
refugee_data$arrivals <- as.numeric(refugee_data$arrivals)
refugee_data$city_pop <- as.numeric(refugee_data$city_pop)
refugee_data$city_pop_avg <- as.numeric(refugee_data$city_pop_avg)
refugee_data$state_pop <- as.numeric(refugee_data$state_pop)
refugee_data$state_pop_avg <- as.numeric(refugee_data$state_pop_avg)
refugee_data$lat <- as.numeric(refugee_data$lat)
refugee_data$lng <- as.numeric(refugee_data$lng)

# Load relevant packages
library(pacman)
p_load(dplyr, knitr, kableExtra, ggplot2, tidyr, scales)





# 1. Function for creating any subset of interest ----

# First, add columns for refugees of nationality X
# (will be helpful later if interested in the most common nationality of refugee arrivals by location)

nationalities <- unique(refugee_data$nationality)

for(i in 1:length(nationalities)) {
  refugee_data <- mutate(refugee_data, total_country = ifelse(nationality==nationalities[i], arrivals, 0))
  colnames(refugee_data)[length(refugee_data)] <- c(paste0(nationalities[i], "_arrivals"))
}

# Overview of subset options

# Refugee arrivals:
# a. Total or of nationality X?
# b. By city or by state?
# c. In what year or years from 2002 to 2019?
# d. As an absolute amount or as a percentage of the city/state population?
# e. What is the most common nationality of refugee arrivals in this location and year?
# f. What is the minimum city population for a city to be eligble to be mapped?
# g. What is the total number of cities when mapping the top N cities by (subset of) refugee arrivals?




# function for creating desired subset

# the function has default values for all arguments

subset_maker <- function(natlty_input = "",         # e.g. "Iran"
                         city_or_state = "City",    # "City" or "State"
                         year_start = 2002,         # year start
                         year_end = year_end,     # year end (if you want multiple years)
                         abs_or_pct = "Abs",        # "Abs" or "Pct": arrivals as absolute number, or
                                                    # as a percentage of the population?
                         most_common_natlty = 1,    # 1 if you want to calc mcn; 0 otherwise
                         city_pop_min = 5000,       # number input if city (useful if we're looking at
                                                    # top N cities of refugees as pct of city pop)
                         number_cities = 100)       # number input if city
  {
  
  subset_data <- refugee_data
  
  # 1. Nationality
  
  # filter by nationality if one is given
  if (nchar(natlty_input) > 0) {
    subset_data <- filter(subset_data, nationality == natlty_input)
  }
  
  
  # 2. City or state
  
  # CITY
  # aggregate data at the state-city-year level
  # if you filtered by nationality, this step will elicit no further aggregation
  # if you did not filter by nationality and are instead looking for total arrivals,
  # this step aggregates across all nationalities
  vars_for_grouping <- c("state", "city", "year", "city_pop", "state_pop", "lat", "lng")
  # include all "arrivals" (total & total from nationality 1,2,..n) as variables to sum
  variables <- colnames(subset_data[,c(5, 12:133)])
  subset_data <- subset_data %<>%group_by(.dots = vars_for_grouping) %>%
    summarise_at(variables, sum)
  # use city_pop as relevant pop variable
  subset_data$pop <- subset_data$city_pop
  # STATE
  # if state, subset further by state; use state_pop as relevant pop variable
  if (city_or_state == "State"){
    state_vars_for_grouping <- c("state", "year", "state_pop")
    subset_data <- subset_data %<>%group_by(.dots = state_vars_for_grouping) %>%
      summarise_at(variables, sum)
    subset_data$pop <- subset_data$state_pop
  }
  
  
  # 3. Year input
  
  # filter by year if only looking at one year
  if (year_start == year_end){
    subset_data <- filter(subset_data, year == year_start)
  }
  
  # if looking at multiple years, subset as follows
  if (year_start < year_end){
    
    # filter down to relevant years
    subset_data <- filter(subset_data, year >= year_start & year <= year_end)
    # add pop as another variable to summarize; will ultimately calculate mean, but
    # first sum like the others
    # recall that the current list includes all "arrivals" columns
    # recall that "pop" in state pop for state subsets and city pop for city subsets
    variables <- c(variables, "pop")
    
    # calculation is summarized at different levels depending on city vs. state decision
    vars_for_grouping <- c("state", "city", "lat", "lng")
    if (city_or_state == "State"){
      vars_for_grouping <- c("state")
    }

    subset_data <- subset_data %<>%group_by(.dots = vars_for_grouping) %>%
      summarise_at(variables, sum)
    # when looking at multiple years, the mean population across these years 
    # (at the geographic level of interest) replaces the annual pop estimate
    subset_data$pop <- round(subset_data$pop / (year_end - year_start + 1), 0)
  }
  
  
  # 4. Absolute or percent
  
  # Create a common variable of interest, "outcome", which captures the relevant outcome for each subset
  # total arrivals, total arrivals by nationality, absolute or percent arrivals...
  
  subset_data$outcome <- subset_data$arrivals
  # if percent is the outcome of interest, calculate as a fraction of the population of interest
  if (abs_or_pct == "Pct"){
    subset_data$outcome <- subset_data$arrivals / subset_data$pop
  }
  subset_data<- subset(subset_data, select=-c(arrivals))
  
  
  # 5. Most common nationality
  
  # Calculate the most common nationality of refugee arrivals at the level you've aggregated to
  # Sometimes, this is uninteresting (if aggregating to Iran arrivals only, Iran will be most common)
  # But it's an interesting / useful metric if looking at all arrivals
  
  if (most_common_natlty == 1){
    a <- which( colnames(subset_data) == "Somalia_arrivals" )
    b <- a + length(nationalities) - 1
    subset_data$most_common_natlty <- colnames(subset_data[a:b])[max.col(subset_data[a:b],
                                                                              ties.method="first")]
    subset_data$most_common_natlty <- gsub("_arrivals", "", subset_data$most_common_natlty)
  }
  
  
  # 6. City population minimum &
  # 7. Number of cities 
  
  subset_data <- subset_data[with(subset_data, order(-outcome)),]
  if (city_or_state == "City"){
    subset_data <- subset(subset_data, subset_data$pop > city_pop_min)
    if (nrow(subset_data) > number_cities) {
      subset_data <- subset_data[1:number_cities,]
    }
  }
  
  
  # Clean and reorder columns for optimal viewing in R
  # These are now superfluous
  if("state_pop" %in% colnames(subset_data)){
    subset_data <- subset(subset_data, select=-c(state_pop))
  }
  if("city_pop" %in% colnames(subset_data)){
    subset_data <- subset(subset_data, select=-c(city_pop))
  }
  # Reordering to get the most interesting info to the front
  a <- which( colnames(subset_data) == "Somalia_arrivals" ) - 1
  refcols <- colnames(subset_data[1:a])
  refcols <- c(refcols, "pop", "outcome")
  subset_data <- subset_data[, c(refcols, setdiff(names(subset_data), refcols))]
  
  # Return data
  return(subset_data)
}





# 2. Function for mapping the data ----

# Download all maps packages that may ultimately prove useful down the line

library(pacman)
p_load(dplyr, tidyverse, data.table, rmarkdown, knitr, tinytex, magrittr,
       gplots,
       viridis, RColorBrewer, scales,
       grDevices, graphics,
       mapproj, ggmap,
       rgdal, tmap, 
       maptools, tmaptools,
       tidycensus, 
       censusapi,
       plotly
)


# Function

data_mapper <- function(df = df){
  
  # default value for the dataframe (df) used is the most recent dataframe named "df"
  # create quantiles for colors in bubble maps
  df$o_ranks <- rank(df$outcome, ties.method = "first")
  df$q <- with(df, cut(o_ranks, quantile(o_ranks)))
  levels(df$q) <- paste(c("1st", "2nd", "3rd", "4th"), "Quantile")
  df$q <- as.ordered(df$q)
  
  # establish US map background
  g <- list(
    scope = 'usa',
    projection = list(type = 'albers usa'),
    showland = TRUE,
    landcolor = toRGB("gray85"),
    subunitwidth = 1,
    countrywidth = 1,
    subunitcolor = toRGB("white"),
    countrycolor = toRGB("white")
  )
  
  # create title
  title_abs_or_pct <- ifelse(abs_or_pct == "Abs", "Total", "Percent")
  title_natlty <- ifelse (natlty_input == "", "", natlty_input)
  title_year <- paste0(year_start)
  if (year_start<year_end){
    title_year <- paste0(year_start, "-", year_end)
  }
  
  # create title, as a function of user data inputs
  title <- paste0("Top Cities by ", 
                  title_abs_or_pct, 
                  " ", 
                  "<br />",
                  title_natlty, 
                  " Refugee Arrivals: ",
                  "<br />",
                  title_year)
  
  # move legend to ideal location
  l <- list(
    x = 1, 
    y = 0.5
  )
  
  # Make outcomes on map read as percentages if relevant
  df$outcome_map <- df$outcome
  if (mean(df$outcome) < 1){
    df$outcome_map <- percent(df$outcome)
  }
  
  # Create plot
  
  p <- plot_geo(df, locationmode = 'USA-states', sizes = c(25, 250)) %>%
    add_markers(
      x = ~lng, y = ~lat, size = ~outcome, color = ~q,
      hoverinfo = "text", text = ~paste(df$city, ",", df$state, ":", "<br />", 
                                        df$outcome_map, "refugees")) %>%
    layout(title = title, geo = g, legend = l)
    
  p
}




# 3. Function for printing a table of the data ----

data_tabler <- function(df = df){
  
  # default value for the dataframe (df) used is the most recent dataframe named "df"
  
  # create title
  title_abs_or_pct <- ifelse(abs_or_pct == "Abs", "Total", "Percent")
  title_natlty <- ifelse (natlty_input == "", "", natlty_input)
  title_year <- paste0(year_start)
  if (year_start<year_end){
    title_year <- paste0(year_start, "-", year_end)
  }
  
  # create title, as a function of user data inputs
  title <- paste0("Top Cities by ", 
                  title_abs_or_pct, 
                  " ",
                  title_natlty, 
                  " Refugee Arrivals: ",
                  title_year)
  
  # Make outcomes on map read as percentages if relevant
  if (mean(df$outcome) < 1){
    df$outcome <- percent(df$outcome)
  }
  
  # Grab relevant columns only
  
  if("city" %in% colnames(df)){
    df_table <- subset(df, select=c(state,city,outcome))
    colnames(df_table) <- c("State", "City", "Arrivals")
  } else{
    df_table <- subset(df, select=c(state,outcome))
    colnames(df_table) <- c("State", "Arrivals")
  }
  
  # Keep top 15 columns only
  
  if(nrow(df_table) > 15){
    df_table <- df_table[1:15,]
  }
  
  # Create table
  
  df_table <- kable(df_table,
                    caption = title,
                    digits = 2, 
                    booktabs=TRUE)
  
  df_table %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), 
                  full_width = F,
                  font_size = 12)
}

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
natlty_input        <- ""
city_or_state       <- "City"
year_start          <- 2002
year_end            <- 2019
abs_or_pct          <- "Abs"
most_common_natlty  <- 1
city_pop_min        <- 5000
number_cities       <- 100

# run subset_maker function with user input data
df <- subset_maker(natlty_input = natlty_input,
                   city_or_state = city_or_state,
                   year_start = year_start,
                   year_end = year_end,
                   abs_or_pct = abs_or_pct,
                   most_common_natlty = most_common_natlty,
                   city_pop_min = city_pop_min,
                   number_cities = number_cities)

# make map
data_mapper(df = df)
```



## Example 1

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# make table
data_tabler(df = df)
```



## Example 2

```{r, echo=FALSE, message=FALSE, warning=FALSE}
natlty_input        <- "Syria"
city_or_state       <- "City"
year_start          <- 2016
year_end            <- 2016
abs_or_pct          <- "Abs"
most_common_natlty  <- 1
city_pop_min        <- 5000
number_cities       <- 100

# run subset_maker function with user input data
df <- subset_maker(natlty_input = natlty_input,
                   city_or_state = city_or_state,
                   year_start = year_start,
                   year_end = year_end,
                   abs_or_pct = abs_or_pct,
                   most_common_natlty = most_common_natlty,
                   city_pop_min = city_pop_min,
                   number_cities = number_cities)

# make map
data_mapper(df = df)
```



## Example 2

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# make table
data_tabler(df = df)
```



## Example 3

```{r, echo=FALSE, message=FALSE, warning=FALSE}
natlty_input        <- "Syria"
city_or_state       <- "City"
year_start          <- 2016
year_end            <- 2016
abs_or_pct          <- "Pct"
most_common_natlty  <- 1
city_pop_min        <- 5000
number_cities       <- 100

# run subset_maker function with user input data
df <- subset_maker(natlty_input = natlty_input,
                   city_or_state = city_or_state,
                   year_start = year_start,
                   year_end = year_end,
                   abs_or_pct = abs_or_pct,
                   most_common_natlty = most_common_natlty,
                   city_pop_min = city_pop_min,
                   number_cities = number_cities)

# make map
data_mapper(df = df)
```



## Example 3

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# make table
data_tabler(df = df)
```



## Example 4

```{r, echo=FALSE, message=FALSE, warning=FALSE}
natlty_input        <- "Cuba"
city_or_state       <- "City"
year_start          <- 2008
year_end            <- 2012
abs_or_pct          <- "Abs"
most_common_natlty  <- 1
city_pop_min        <- 5000
number_cities       <- 100

# run subset_maker function with user input data
df <- subset_maker(natlty_input = natlty_input,
                   city_or_state = city_or_state,
                   year_start = year_start,
                   year_end = year_end,
                   abs_or_pct = abs_or_pct,
                   most_common_natlty = most_common_natlty,
                   city_pop_min = city_pop_min,
                   number_cities = number_cities)

# make map
data_mapper(df = df)
```



## Example 4

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# make table
data_tabler(df = df)
```



## Example 5

```{r, echo=FALSE, message=FALSE, warning=FALSE}
natlty_input        <- "Somalia"
city_or_state       <- "City"
year_start          <- 2008
year_end            <- 2012
abs_or_pct          <- "Abs"
most_common_natlty  <- 1
city_pop_min        <- 5000
number_cities       <- 100

# run subset_maker function with user input data
df <- subset_maker(natlty_input = natlty_input,
                   city_or_state = city_or_state,
                   year_start = year_start,
                   year_end = year_end,
                   abs_or_pct = abs_or_pct,
                   most_common_natlty = most_common_natlty,
                   city_pop_min = city_pop_min,
                   number_cities = number_cities)

# make map
data_mapper(df = df)
```



## Example 5

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# make table
data_tabler(df = df)
```



## Example 6

```{r, echo=FALSE, message=FALSE, warning=FALSE}
natlty_input        <- "Iran"
city_or_state       <- "City"
year_start          <- 2008
year_end            <- 2012
abs_or_pct          <- "Abs"
most_common_natlty  <- 1
city_pop_min        <- 5000
number_cities       <- 100

# run subset_maker function with user input data
df <- subset_maker(natlty_input = natlty_input,
                   city_or_state = city_or_state,
                   year_start = year_start,
                   year_end = year_end,
                   abs_or_pct = abs_or_pct,
                   most_common_natlty = most_common_natlty,
                   city_pop_min = city_pop_min,
                   number_cities = number_cities)

# make map
data_mapper(df = df)
```



## Example 6

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# make table
data_tabler(df = df)
```



## Instructions for making your own maps and tables

>- Locate code at my github page: https://github.com/matt-brundage/239T_final/blob/master/Code/data_mapping.R

>- Run the entire code

>- Navigate to the final section of the code, entitled "4. USER INPUT for City Bubble Map and Table"

>- Follow instructions to input relevant variables

>- Run the data_mapper and data_tabler functions immediately below



## Instructions for making your own maps and tables

The code and instructions in data_mapping.R read as follows.

```{r}
# 4. USER INPUT for City Bubble Map and Table ----

# Conceptual overview of variables

# a. Total or of nationality X?
# b. By city or by state?
# c. In what year or years from 2002 to 2019?
# d. As an absolute amount or as a percentage of the city/state population?
# e. What is the most common nationality of refugee arrivals in this location and year?
# f. What is the minimum city population for a city to be eligble to be mapped?
# g. What is the total number of cities when mapping the top N cities by (subset of) refugee arrivals?
```



## Instructions for making your own maps and tables

```{r}
# Variable input instructions

# natlty_input: Enter national, e.g. "Iran." Enter "" if total refugees, irrespective of nationality
# city_or_state: Enter "City" or "State." For Bubble maps, only "City" will work.
# year_start: Enter any year from 2002 to 2019.
# year_end: Enter any year from 2002 to 2019. Can be the same as year_start if you only want one year.
# abs_or_pct: Enter "Abs" or "Pct". Arrivals as absolute number, or as a percentage of the population
# most_common_natlty: 1 if this is of interest; 0 otherwise
# city_pop_min: number input
# number_cities: number input

# once you have completed user input, run R from here down
# example input shown here

natlty_input        <- "Syria"
city_or_state       <- "City"
year_start          <- 2016
year_end            <- 2016
abs_or_pct          <- "Abs"
most_common_natlty  <- 1
city_pop_min        <- 5000
number_cities       <- 100
```



## Instructions for making your own maps and tables

Of course, if you are interesting in seeing any particular visualization and would rather not go through the process of making your own table, just shoot me an email and I should be able to get any visualizations to you within 24 hours

---